Formalization of Abstract Rewriting

### 2025-02-23 Relations/Paper/Definitions.tex
Figure for WF

Having a size function that maps to the natural numbers is computationally taxing to satisfy. 
Further, it takes us outside of purely relational theory. The following definitions are for properties which allow us to forgo $\Inc$, for further discussion see Subsection \ref{subsec:def}.
\begin{definition} \label{def:rp} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description} 
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a$ is an $s$-bound] if $s (i) \mstep a$ for all $i \in \nat$. 
        \item[$a \in \BP_R$] if all $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound.  
        \item[$R \models \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.

        \item[$R \models \RP$] \emph{$R$ has the recurrence property} if every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that 
        $s (i) \mstep x \implies x \mstep s (i)$ for all $x\in A$. See Definition \ref{def:mf}. 
    \end{description}
\end{definition}
In \terese $\mathsf{Ind}$ is used to denote $\BP$.

>>>>

Having a size function that maps to the natural numbers is computationally taxing to satisfy. 
Further, it takes us outside of purely relational theory. The following definitions are for properties which allow us to forgo $\Inc$, for further discussion see Subsection \ref{subsec:def}.
\begin{definition} \label{def:rp} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description} 
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a$ is an $s$-bound] if $s (i) \mstep a$ for all $i \in \nat$. 
        \item[$a \in \BP_R$] if all $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound.  
        \item[$R \models \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.
        \item[$R \models \RP$] \emph{$R$ has the recurrence property} if for every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that 
        $s (i) \mstep x \implies x \mstep s (i)$ for all $x\in A$. See Definition \ref{def:mf}. 
    \end{description}
\end{definition}
In \terese $\mathsf{Ind}$ is used to denote $\BP$.

### 2025-02-23 Relations/Paper/Formalization.tex
Figure for WF

As already mentioned, the $\Inc$ property was one we wanted to move away from. We saw that $\Inc$ was being used only in Theorem 1.2.3 
and that the crucial facet of $\Inc$ for the proof of the theorem was that no $R$-increasing 
sequence can have an upper bound.
We established that the properties $\RP$ and $\RPm$ were sufficient for the proving the theorems. This in turn led us to 
examine the condition in $\RP$ more closely. This condition provides a generalization of the notion of normal form, 
which we analyze in detail in Section \ref{sec:Implications}. \sacomment{Try and get this para to flow better.}

@@

\subsection{Theorems}
\subsubsection{Newman's Lemma}
There are three proofs of Newman's Lemma ($\SN \land \WCR \implies \CR$) given in \terese. 
The first and third of these proofs make use of classical properties. The second proof is amenable to a 
constructive formalization as shown in the following function in \texttt{ARS-NewmansLemma.agda}:

@@

There are four parts to this theorem. 
\begin{description}
    \item[($i$)] $\gWN \and \gUN \implies \gBP$
    \item[($ii$)] $\gBP \land \gInc \implies \gSN$  
    \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item[($iv$)] $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
\end{description}
($i$) is formalized in: 
\begin{itemize}
    \item \verb|i : R isWN → R isUN → R isBP|
    \item \verb|i→ : R isWN → R isUN→ → R isBP|
\end{itemize}
In $\mathtt{i\to}$ we again provide a generalization by using $\gUNto$ rather than $\gUN$.

A near formalization of ($ii$) is provided in: 
\begin{itemize}
    \item \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|
\end{itemize}
\sacomment{Explain here why have we provided a near formalization rather than sticking to the same properties}

($iii$) is formalized in: 
\begin{itemize}
    \item \verb|iii : R isWN → R isWCR → R isRP- → dec (SN) → R isSN|
\end{itemize}
As can be seen we have substituted $\gInc$ with $\gRPm$ which generalizes this theorem. We have also included the assumption that $\gSN$ 
is decidable \sacomment{Is decidability being discussed elsewhere in the paper that we can reference here.} as this property enables us 
to construct a witness for an element that either does or does not have the property $\SN$. Such a witness was necessary for our proof. \sacomment{More to say on this theorem?}

\sacomment{neeed to define CP}
The reverse implication of ($iv$) is formalized in: 
\begin{itemize}
    \item \verb|iv : R isCP → R isCR|
\end{itemize}
We did not formalize $\gCR \to gCP$ as it requires the ARS to be countable \sacomment{More to say?}.

>>>>

As already mentioned, the $\Inc$ property has undesired requirements. $\Inc$ is used only in Theorem 1.2.3 
and the crucial facet of $\Inc$ for the proof of the theorem is that no $R$-increasing 
sequence can have an upper bound.
We established that the properties $\RP$ and $\RPm$ were sufficient for proving the theorem (see Subsection \ref{subsec:theorems}). This in turn led us to 
examine the condition in $\RP$ more closely. This condition provides a generalization of the notion of normal form, 
which we analyze in detail in Section \ref{sec:Implications}.

@@

\subsection{Theorems}\label{subsec:theorems}
\subsubsection{Newman's Lemma}
There are three proofs of Newman's Lemma ($\SN \land \WCR \implies \CR$) given in \terese. 
The first and third of these proofs make use of classical properties. The second proof is amenable to a 
constructive formalization as shown in the following function in \texttt{ARS-NewmansLemma.agda}:

@@

There are four parts to this theorem. 
\begin{description}
    \item[($i$)] $\gWN \land \gUN \implies \gBP$
    \item[($ii$)] $\gBP \land \gInc \implies \gSN$  
    \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item[($iv$)] $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
\end{description}
($i$) is formalized in: 
\begin{itemize}
    \item \verb|i : R isWN → R isUN → R isBP|
    \item \verb|i→ : R isWN → R isUN→ → R isBP|
\end{itemize}
In $\mathtt{i\to}$ we again provide a generalization by using $\gUNto$ rather than $\gUN$.

A formalization of ($ii$) is provided in: 
\begin{itemize}
    \item \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|
\end{itemize}
This is a fundamentally classical result, and so the result is only classically equivalent to $\gSN$.
\sacomment{Double check the above explanation.}

($iii$) is formalized in: 
\begin{itemize}
    \item \verb|iii : R isWN → R isWCR → R isRP- → dec (SN) → R isSN|
\end{itemize}
As can be seen we have substituted $\gInc$ with $\gRPm$ which generalizes this theorem. We have also included the assumption that $\gSN$ 
is decidable as this property enables us 
to construct a witness for an element that either does or does not have the property $\SN$. Such a witness was necessary for our proof. \sacomment{More to say on this theorem?}

The reverse implication of ($iv$) is formalized in: 
\begin{itemize}
    \item \verb|iv : R isCP → R isCR|
\end{itemize}
We did not formalize $\gCR \to \gCP$ as it requires the ARS to be countable \sacomment{Add this proof to future todo}.

### 2025-02-23 Relations/Paper/Implications.tex
Figure for WF

% From the perspective of programming language theory, the two properties of abstract rewriting 
% systems that are most interesting are $\SN$ and $\CR$. The following hierarchies 
% show chains of implications starting from these properties.
As discussed in Subsection \ref{subsec:def} we put forward $\RP$ as an alternative for $\Inc$. It is through examining $\RP$ that 
we came to the below definitions.

\begin{definition}\label{def:mf}  Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \MF_R$] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[$a \in \WM_R$] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[$a \in \SM_R$]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is also strongly minimalizing.
        \item[$a \in \MP_R$] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.

@@

\sacomment{Need to relate SM and SMseq. Probably define SMseq and mention 
we can only prove this for SMseq.}
\begin{proposition}\label{prop:SMRP}
    $SM \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

Trivially, the following also holds. 
\begin{proposition}\label{prop:nftomf}
    $a\in NF_R \implies a \in MF_R$ 
\end{proposition}    
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{ARS-Implications.agda}.
\end{proof}

@@

\sacomment{Are we going to define this property here or elsewhere?}

The isMinDec property states that all elements either have a relation to some other element, or 
there is no other element to which they have a relation. If the element relates to no other element then it is terminating 
and so trivially $\WN$. If the element does relate to some other element $y \in A$, then we recursively apply our claim to 
$y$ and because we know that we have the property of $\SN$ we must eventually reach a terminating element.

@@

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\begin{table}[h!]
    \centering
    \caption{Local implications}
    \begin{tabular}{|>{\columncolor{gray!30}}l|c|c|c|c|c|}
    \hline
    \rowcolor{gray!30}     & $\WM$         & $\WN$         & $\SM$             & $\SMandWN$        & $\SN$ \\
    \hline
    $\UNto$ &  CE-\ref{CE:4}, CE-\ref{CE:12}     & CE-\ref{CE:4}, CE-\ref{CE:7}       & $\CR$\footnotemark[1] , CE-\ref{CE:9}          & $\CR$\footnotemark[1] , CE-\ref{CE:4}            & $\CR$\footnotemark[1] , $\SN$ \\
    \hline
    $\NP$ & CE-\ref{CE:10},CE-\ref{CE:9}      & $\CR$, CE-\ref{CE:7}    & $\CR$\footnotemark[1]  , CE-\ref{CE:9}          & $\CR$, $\SN$      & $\CR$\footnotemark[2] , $\SN$ \\
    \hline
    $\MP$ & $\CR$ , CE-\ref{CE:9}    & $\CR$, CE-\ref{CE:7}    & $\CR$\footnotemark[2] , CE-\ref{CE:9} & $\CR$, $\SN$      & $\CR$\footnotemark[2] , $\SN$ \\
    \hline
    $\CR$   & $\CR$, CE-\ref{CE:9}     & $\CR$, CE-\ref{CE:7}     & $\CR$, CE-\ref{CE:9}        & $\CR$, $\SN$     & $\CR$ , $\SN$ \\
    \hline

    \end{tabular}
\end{table}
\begin{table}[h!]
    \centering
    \caption{Global implications}
    \begin{tabular}{|>{\columncolor{gray!30}}l|c|c|c|c|c|}
    \hline
    \rowcolor{gray!30}     & $\gWM$         & $\gWN$         & $\gSM$             & $\gSMandWN$        & $\gSN$ \\
    \hline
    $\gUNto$ &  CE-\ref{CE:2}, CE-\ref{CE:9}     & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[1] , CE-\ref{CE:9}          & $\gCR$, $\gSN$      & $\gCR$\footnotemark[2] , $\gSN$ \\
    \hline
    $\gNP$ & CE-\ref{CE:3} , CE-\ref{CE:9}      & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[1] , CE-\ref{CE:9}          & $\gCR$, $\gSN$      & $\gCR$\footnotemark[2] , $\gSN$ \\
    \hline
    $\gMP$ & $\gCR$ , CE-\ref{CE:9}    & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[2] , CE-\ref{CE:9} & $\gCR$, $\gSN$    & $\gCR$\footnotemark[2] , $\gSN$ \\
    \hline
    $\gCR$   & $\gCR$, CE-\ref{CE:9}     & $\gCR$, CE-\ref{CE:7}     & $\gCR$, CE-\ref{CE:9}        & $\gCR$, $\gSN$      & $\gCR$ , $\gSN$ \\
    \hline

    \end{tabular}
\end{table}
\footnotetext[1]{This implication also requires $\gWCR$.}
\footnotetext[2]{This implication also requires either $\gWCR$ or the classical property required to go from 
$\SN \to \WN$ or $\SM \to \WM$.}
\renewcommand*{\thefootnote}{\arabic{footnote}}

@@

@@

% The tables show a number of changes when moving from properties holding locally to globally. 
% Globally $\WN \land \UNto \implies \CR$ holds whereas it does not locally. Similarly, $\SM \land \WN \implies \SN$ holds 
% globally but not locally. 
\subsection{Relation between $\SM$, $\WN$, and $\SN$}
One key takeaway from the tables is that we obtain completeness when we have the properties $\SM$, $\WN$, and $\WN$. We also 
obtain completeness when we have $\gWN$ and $\gSM$. 
To show that $\gSM \land \gWN \implies \gSN$ we built on our proof that $\SM \land \WN \land \NP \implies \SN$. If $\gWN$ holds 
then our proof no longer requires the property $\NP$. This progression of proofs can be seen in the functions:

@@

\sacomment{Are there any other features of the table that we would like to draw out?}
\sacomment{We should remember that If global WCR and local SM implies local CR, then if we also have local WN, we get local SN.}

>>>>

% From the perspective of programming language theory, the two properties of abstract rewriting 
% systems that are most interesting are $\SN$ and $\CR$. The following hierarchies 
% show chains of implications starting from these properties.
As discussed in Subsection \ref{subsec:def} we put forward $\RP$ as an alternative property for $\Inc$. It is through examining $\RP$ that 
we came to the definitions below.

\begin{definition}\label{def:mf}  Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \MF_R$] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[$a \in \WM_R$] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[$a \in \SMseq_R$] \emph{$a$ is sequentially strongly minimalizing} if for every $R$-increasing sequence $s$ starting at $a$, there exists 
        an $i \in \nat$ such that $s (i) \in \MF$. 
        \item[$a \in \SM_R$]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is also strongly minimalizing.
        \item[$a \in \MP_R$] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.

@@

We show that $\SM \implies \SMseq$, however $\SMseq$ is only classically equivalent to $\SM$. The 
following proposition relates these new definitions with our previous $\RP$ definitions.
\begin{proposition}\label{prop:SMRP}
    $SMseq \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

Trivially, the following also holds. 
\begin{proposition}\label{prop:nftomf}
    $\NF \implies \MF$ 
\end{proposition}    
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{ARS-Implications.agda}.
\end{proof}

@@

The isMinDec property \sacomment{refer to list of classical properties here} states that all elements either have a relation to some other element, or 
there is no other element to which they have a relation. If the element relates to no other element then it is terminating 
and so trivially $\WN$. \greyout{If the element does relate to some other element $y \in A$, then we recursively apply our claim to 
$y$ and because we know that we have the property of $\SN$ we must eventually reach a terminating element.}

@@

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\begin{table}[h!]
    \centering
    \begin{tabular}{|>{\columncolor{gray!30}}l|c|c|c|c|c|}
        \hline
        \rowcolor{gray!30}     & $\WM$         & $\WN$         & $\SM$             & $\SMandWN$        & $\SN$ \\
        \hline
        $\UNto$ &  CE-\ref{CE:4}, CE-\ref{CE:12}     & CE-\ref{CE:4}, CE-\ref{CE:7}       & $\CR$\footnotemark[1] , CE-\ref{CE:9}          & $\CR$\footnotemark[1] , CE-\ref{CE:4}            & $\CR$\footnotemark[1] , $\SN$ \\
        \hline
        $\NP$ & CE-\ref{CE:10},CE-\ref{CE:9}      & $\CR$, CE-\ref{CE:7}    & $\CR$\footnotemark[1]  , CE-\ref{CE:9}          & $\CR$, $\SN$      & $\CR$\footnotemark[2] , $\SN$ \\
        \hline
        $\MP$ & $\CR$ , CE-\ref{CE:9}    & $\CR$, CE-\ref{CE:7}    & $\CR$\footnotemark[2] , CE-\ref{CE:9} & $\CR$, $\SN$      & $\CR$\footnotemark[2] , $\SN$ \\
        \hline
        $\CR$   & $\CR$, CE-\ref{CE:9}     & $\CR$, CE-\ref{CE:7}     & $\CR$, CE-\ref{CE:9}        & $\CR$, $\SN$     & $\CR$ , $\SN$ \\
        \hline

    \end{tabular}
    \caption{Local implications}
\end{table} 
\vspace{-1cm}
% \\[-0.5cm]
\begin{table}[h!]
    \centering
    \begin{tabular}{|>{\columncolor{gray!30}}l|c|c|c|c|c|}
        \hline
        \rowcolor{gray!30}     & $\gWM$         & $\gWN$         & $\gSM$             & $\gSMandWN$        & $\gSN$ \\
        \hline
        $\gUNto$ &  CE-\ref{CE:2}, CE-\ref{CE:9}     & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[1] , CE-\ref{CE:9}          & $\gCR$, $\gSN$      & $\gCR$\footnotemark[2] , $\gSN$ \\
        \hline
        $\gNP$ & CE-\ref{CE:3} , CE-\ref{CE:9}      & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[1] , CE-\ref{CE:9}          & $\gCR$, $\gSN$      & $\gCR$\footnotemark[2] , $\gSN$ \\
        \hline
        $\gMP$ & $\gCR$ , CE-\ref{CE:9}    & $\gCR$, CE-\ref{CE:7}    & $\gCR$\footnotemark[2] , CE-\ref{CE:9} & $\gCR$, $\gSN$    & $\gCR$\footnotemark[2] , $\gSN$ \\
        \hline
        $\gCR$   & $\gCR$, CE-\ref{CE:9}     & $\gCR$, CE-\ref{CE:7}     & $\gCR$, CE-\ref{CE:9}        & $\gCR$, $\gSN$      & $\gCR$ , $\gSN$ \\
        \hline

    \end{tabular}
    \caption{Global implications}
\end{table}
\footnotetext[1]{This implication also requires $\gWCR$.}
\footnotetext[2]{This implication also requires either $\gWCR$ or the classical property required to go from 
$\SN \to \WN$ or $\SM \to \WM$.}
\renewcommand*{\thefootnote}{\arabic{footnote}}

@@

\begin{remark} The following are the key takeaways from the implication tables. \hfill
    \begin{enumerate}
        \item The property $\gWCR$ can sometimes be a substitute for the classical 
        property required for $\SN \implies \WN$ (and similarly $\SM \implies WM$).   
        \item $\gWN \land \gUNto \implies \gCR$ doesn't require any classical assumptions
         or $\gWCR$ for $\gCR$, unlike $\gSN \land \gUNto$. 
        \item $\gWN \land \gUNto \implies \gCR$ only holds when the properties are global. The 
        counterexample provided illustrates why $\WCR \land \WN \land \UNto \nRightarrow \CR$.
        \item Both $\SM$ and $\WN$ are required for $\SN$, this connection is explored in Subsection \ref{subsec:SMWNSN}.
        \item We have a generalization of Newman's Lemma as $\SM \land \gWCR \implies \CR$. This is explored in Subsection \ref{subsec:newnewman}.
    \end{enumerate}
\end{remark}

@@

% The tables show a number of changes when moving from properties holding locally to globally. 
% Globally $\WN \land \UNto \implies \CR$ holds whereas it does not locally. Similarly, $\SM \land \WN \implies \SN$ holds 
% globally but not locally. 
\subsection{Relation between $\SM$, $\WN$, and $\SN$} \label{subsec:SMWNSN}
One key takeaway from the tables is that we obtain completeness when we have the properties $\SM$, $\WN$, and $\NP$. We also 
obtain completeness when we have $\gWN$ and $\gSM$. 
To show that $\gSM \land \gWN \implies \gSN$ we built on our proof that $\SM \land \WN \land \NP \implies \SN$. If $\gWN$ holds 
then our proof no longer requires the property $\NP$. This progression of proofs can be seen in the functions:

@@
